A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-20 latex/main.tex
Various updates suggested through ChatGPT discussions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%  Welcome to the Cell Press LaTeX template,
%%%  version 1.10. This is a minimalist template
%%%  to help you organize your article for
%%%  publication at Cell Press. PLEASE NOTE:

%%%  (1) If you submit your final manuscript materials
%%%  in LaTeX format, our typesetters will prepare
%%%  a Word file for use in production. This conversion
%%%  process allows us to copyedit your paper, fix
%%%  any typos, and add formatting and tagging. The
%%%  conversion process will add approximately 3
%%%  business days to the production timeline.
%%%  Authors using LaTeX should keep this in mind
%%%  when considering deadlines.

%%%  (2) Keep your LaTeX files as simple as possible.
%%%  Avoid the use of elaborate local macros and/or
%%%  customized external style files. If you need
%%%  additional macros, please keep them simple and
%%%  include them in the actual .tex document preamble.
%%%  Source code should be set up so that all .sty
%%%  and .bst files called by the main .tex file are
%%%  in the same directory as the main .tex file.

%%%  (3) Cell Press publishes more than 40 journals,
%%%  some of which may have different or additional
%%%  formatting requirements not specified in this
%%%  template. When revising your paper before
%%%  acceptance, please review the formatting
%%%  guidelines, including the Final Files Requirements,
%%%  for the journal you are publishing with.

%%%  Please send feedback on this template to lshipp@cell.com.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

[NOTE: the discussion needs to be reworked, and hasn't been updated
to account for changes. See the new draft subsection on prevalence of recombination,
though]

@@

We present sc2ts, a real-time ARG inference method,
and applied it to build a $\sim$2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
By identifying specific samples as parents,
sc2ts enables more precise interpretation of recombinants
than state-of-the-art recombination detection methods,
which find recombinants by matching a given sample
to pre-defined population-averaged mutational profiles of Pango lineages.
Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution,
in contrast to RIPPLES \cite{Turakhia2022}
which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution.
Combined with new methods and tools to evaluate evidence for recombination,
sc2ts is a significant technological advance over current methods
for the discovery and analysis of recombinants and
efficient inference of SARS-CoV-2 genealogy.

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies
using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence
with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data.
In that space, the availability of ARGs reconstructed from large datasets
has already had tremendous impact on improving our understanding of
human genetic ancestry and past demography \cite{Wohns2022}.
We expect similar leaps to be enabled
through tailoring ARG reconstruction methods
to data from other species and pathogens.
Specifically, existing inference methods can be adapted
to use the sc2ts ARG for the genealogy-based study of selection \cite{hejase2022deep}
or the geographic spread of the virus through time \cite{Grundler2025geographic}.
This is in addition to simpler analyses,
such as calculating mutation spectra
or querying the number or frequency of mutations at a given position,
which the ARG makes significantly more efficient, and
ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save significant time
and effort for other researchers and enable immediate focus on analyzing
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).

We contribute several novel methods and concepts
which may find applications beyond sc2ts.
First, we devise a simplified LS model that extends the classic concept of parsimony.
In sc2ts, this model allows for accelerated HMM likelihood calculations,
enabling reconstruction of the pandemic-scale ARG presented here.
Existing HMMs used in statistical genetics may be modified
in a similar way for significant speedup.
Second, we propose the concept of mutations averted by recombination,
which leads to an interpretable quantity to weigh the support for 
the hypothesis that a given sample is a recombinant against
the hypothesis that it is a non-recombinant,
considering the sampled genetic diversity in an ARG and a specific HMM.
Third, we propose the concept of HMM cost
to identify poorly explained samples
which are either too divergent or
have a high number of mutations inconsistent with their collection dates.
Here we use HMM cost to reject time-traveling samples,
which no existing phylogenetic method detects and handles during inference
(e.g., in the Viridian UShER phylogeny, parent and child nodes
do not necessarily follow time order \cite{Hunt2024}),
resulting in a cleaner dataset without samples
with wrong collection dates that may affect downstream analyses.
Future work may be to estimate dates
for the suspected time traveling samples using the ARG provided here and
the sc2ts sample matching engine.

Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
but has been understudied in part
due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
While the methods and tools developed here are tailored for SARS-CoV-2,
they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is that
they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops.
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
making them highly adaptable for a broad range of applications.
Together with the pandemic-scale ARG,
we hope that sc2ts will be highly accessible to many researchers,
enabling and accelerating studies of SARS-CoV-2 for years to come.

\subsection*{Prevalance of recombination in SARS-CoV-2}
[Rough first draft taken from earlier location. Sticking in a heading to
make it clear it's distinct from other paragraphs.]

The number of recombination events that we detect is largely 
driven by the choice of recombination penalty, $k$ (or $p$
for RIPPLES). 
At $k=4$ we detect fewer recombinants overall than the combination of 
UShER and RIPPLES at the same threshold. 
However, most of the extra recombinants detected by UShER+RIPPLES
involve violations of time ordering; whether this is due to incorrect
metadata or these are false positives would require further study.
However, at $p=4$, UShER+RIPPLES detects very few of the well-characterised
Pango X-lineages which are detected by sc2ts. At the more lenient
threshold of $p=3$, more of these events are detected (but still fewer
than sc2ts) but at the cost of nearly 4 times more recombination
events overall. Our detailed analysis shows that RIPPLES has a well
defined task, which it performs well, finding the potential events
in an exhaustive search. However, it can only operate within the
input that it is provided, and it is the UShER tree provided with
the Viridian data that does not have clear signals for recombination
of X lineages such as XBB at the $p=4$ threshold. 
The inference of recombination is therefore intrinsically linked
with the underlying genealogy inference method.

[Same point for us with XB. We don't pick up XB because of a long 
branch and we don't have the right parents present.]

[Whether k=3 or k=4 is ``correct'' is a matter for debate, and feeds
into the next point.]

The estimate for effective recombination in SARS-CoV-2 that we obtain
at k=4 is much lower 
than previous studies have suggbalanced
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
In an early analysis, 225 potentially recombinant sequences out 
of 87,695 samples collected before October 2020\cite{Varabyou2021rapid}.
Over the same time period, XX events are present in the sc2ts ARG,
representing a total of XX samples.
In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
589 unique recombination events in a tree of 1.6 million samples
up to May 2021 were identified after extensive quality control.
These nodes had a combined total of 43,104 descendants, suggesting
that approximately 2.7\% of genomes were of recombinant origin.
Over the same time period, the sc2ts ARG contains 14 recombinants
passing QC with a total of 172 descendants. The ARG contains a total
501,235 samples over this time period, giving an overall total fraction of
0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
events and 79-fold descendant samples over the same time period.
The numbers are not directly comparable as the published RIPPLES 
estimate is based on GISAID data which is enriched for
artefacts~\cite{Hunt2024}, but given the observations in 
the previous section on running RIPPLES on the Viridian data
it seems likely that RIPPLES is overestimating the prevalence
of recombination in SARS-CoV-2.

>>>>

% Notes on format:
% The total word count of an article must be under 7,000 (including main figure
% legends but excluding STAR Methods text, supplemental item legends, and
% references section), and there should be no more than seven figures and/or
% tables.

@@

% \section*{DISCUSSION}

% NOTES: paragraph structure. Lead with the biological insights first 
% to prevent the ``this is a methods paper'' criticism

% \begin{itemize}
% \item We present a landmark ARG, which has a comprehensive representation
% of evolutionary history.
% \item This is a conservative lower-bound on the occurance of recombination.
% Very high validation rate post-QC.
% Substantially lower than previous estimate. 
% The first order term of this 
% is the choice of $p$ parameter. With p=4 we show that the ARG contains
% more of the previously identified recombinants.
% \item Truth data for recombination is difficult to find. The closest we have is the 
% pango X lineages. No methods agrees with all of the designations being 
% true recombinants. Our analysis suggests that there is significant over-naming
% with the same underlying recombination event getting associated with 
% multiple X lienages at least twice. Also many lineages are not suggested 
% as being recombinant. This resolves long-standing complexity in 
% the Omicron evolution and simplifies our understanding.
% \item Based on this cleaned up and conservative estimate of recombination
% prevalance. Supports the idea that  recombination frequency in SARS-CoV-2 
% is driven jointly by viral
% load in the population and lineage diversity, implying predictable
% recombination ``hot periods'' during pandemics
% \item Also found evidence that breakpoints are enriched near spike, 
% supporting findings of earlier studies.
% Suggests functional constraints or template-switching
% biases that may shape SARS-CoV-2 evolution.
% Spike gene junctions are recurrent recombination zones, likely under both
% mechanistic and selective pressures.
% \item This ARG is base on a new method, sc2ts. Major methodological 
% breakthrough in terms of handling recombination. We validate the method
% by using results from the literature. We don't use simulations because 
% there's no real point in this case. We're just focusing on SARS-CoV-2.
% We could create some simulations and generate some metrics about how
% well sc2ts picks up simulated recombination events on clean data, 
% but what would that tell us ultimately? (Similarly for tree reconstruction.)
% Certainly standardised simulations would help to refine methods, but
% we have moved the bar forwards massively here and made a great starting
% point for subsequent method development.
% \item There is great potential for using this ARG in future studies,
% and using sc2ts as a baseline to judge future methods.
% \item The tskit and VCF Zarr toolchains enable this research.
% \item We have shown that recombination can be handled, provided
% feasible methods to infer it, and shown that a preexisting tool
% chain can provide quick insignts. The future is bright.

% \end{itemize}

@@

We present here the first pandemic-scale Ancestral Recombination Graph (ARG)
for SARS-CoV-2, reconstructed from 2.48 million genomes. This resource provides
a comprehensive, recombination-aware view of the evolutionary history of the
virus, capturing not only point mutations but also deletions and recombination
events in a single coherent framework. By representing SARS-CoV-2 evolution as
a graph rather than a tree, the ARG unifies the relationships among lineages
that could not previously be reconciled by tree-based phylogenies. The
resulting structure serves as a long-term reference for studies of viral
evolution and as a foundation for future recombination-aware surveillance.

Our inference of recombination represents a conservative lower bound on the
true frequency of recombination in SARS-CoV-2. This conservatism arises from
stringent quality control criteria, including the requirement for multiple
supporting loci for each parental assignment. Even under these strict
thresholds, the ARG contains hundreds of high-confidence recombination events
with a validation rate exceeding 90\% among events passing QC filters. The
overall prevalence of recombination inferred here is substantially lower than
early estimates based on smaller or less curated datasets, but this is
consistent with our aim to prioritise specificity over sensitivity.
Importantly, at the default parsimony penalty ($p = 4$), the ARG recovers the
majority of previously recognised recombinant lineages, including those most
relevant to global transmission dynamics.

Comparisons with the Pango X designations reveal that several nominally
distinct recombinant lineages share common ancestral recombination events. This
over-splitting has led to an inflated perception of recombinant diversity.
Within Omicron, for example, multiple X lineages are better explained as
descendants of a single ancestral recombinant rather than as independent
crossover events. Resolving these redundancies clarifies long-standing
complexity in the Omicron genealogy and simplifies the interpretation of
SARS-CoV-2’s reticulate evolution. Conversely, several designated X lineages
show little evidence of recombination in the ARG, emphasising that not all
phylogenetically unusual lineages are true recombinants. Together, these
results provide a more parsimonious and internally consistent model of
recombinant evolution.

Using the dated ARG, we examined the temporal and epidemiological correlates of
recombination. The frequency of inferred recombination events correlates
strongly with both global case numbers and the diversity of circulating
lineages. This pattern supports the hypothesis that recombination is driven
jointly by viral load in the population, which increases opportunities for
co-infection, and by lineage diversity, which provides distinct parental
genomes. These findings imply predictable ``hot periods'' for recombinant
emergence during pandemics, particularly when multiple antigenically divergent
variants co-circulate at high prevalence. Understanding these dynamics may aid
future outbreak forecasting and genomic surveillance strategies.

Analysis of genomic breakpoint locations revealed enrichment of recombination
near the boundaries of the Spike gene, consistent with previous observations
from smaller datasets. The repeated occurrence of breakpoints in these regions
suggests that structural or functional constraints may influence where viable
recombinants can arise. Mechanistic biases in template switching, combined with
selective pressures on Spike for receptor binding and immune escape, may
together shape the landscape of recombination. The ARG thus not only captures
recombination events but also points to underlying biological mechanisms worthy
of experimental follow-up.

Beyond its biological insights, this study represents a methodological advance
in real-time, recombination-aware phylogenetic inference. The \texttt{sc2ts}
method integrates an efficient, simplified Li--Stephens model with incremental
sample addition, allowing scalable ARG construction from millions of genomes.
We validated the method against empirical data and literature-curated
recombinant lineages rather than simulated datasets, because the global
SARS-CoV-2 collection offers an unparalleled real-world benchmark.
Nevertheless, standardised simulations will remain valuable for refining models
and comparing future implementations. Our results demonstrate that the
\texttt{sc2ts} framework captures the essential genealogical structure of
SARS-CoV-2 while accommodating recombination in a computationally tractable
way.

The ARG produced here is embedded within the \texttt{tskit} and
\texttt{VCF-Zarr} ecosystems, enabling compact storage, rapid analysis, and
open sharing. This integration ensures that the resource is immediately usable
by the broader community for tasks such as mapping selective sweeps, tracing
recombinant origins, and evaluating lineage dynamics. The dataset and
accompanying workflows provide a transparent foundation for benchmarking future
methods and for extending ARG-based analyses to other pathogens.

In summary, this work demonstrates that recombination in SARS-CoV-2 can be
robustly inferred and represented at scale, transforming how we study the
evolution of rapidly diversifying viruses. The ARG provides a unified,
validated view of SARS-CoV-2 evolution that resolves the relationships among
recombinant lineages, links recombination frequency to epidemiological
dynamics, and highlights mechanistic constraints on recombination. More
broadly, it establishes that scalable ARG inference is now a practical and
powerful tool for pandemic genomics.

\subsection*{OUTLOOK}

The pandemic-scale ARG presented here establishes a new foundation for studying
viral evolution in real time. Beyond SARS-CoV-2, the \texttt{sc2ts} framework
can be directly adapted to other recombining pathogens such as influenza, HIV,
and rotaviruses, where complex genealogies have long limited evolutionary
analysis. Integrating the ARG with epidemiological and immunological data will
enable quantitative studies of how recombination interacts with selection,
immune escape, and transmission dynamics. As sequencing becomes increasingly
routine in global surveillance, efficient ARG inference will allow real-time
tracking of recombinant emergence and fitness, supporting faster risk
assessment and lineage designation. The combination of scalable inference, open
data standards, and community tools transforms recombination from a nuisance
factor into an analyzable dimension of viral evolution. We anticipate that this
resource will not only clarify the history of SARS-CoV-2 but also guide the
design of next-generation computational and experimental approaches to
understanding recombination-driven adaptation in RNA viruses.

% [NOTE: the discussion needs to be reworked, and hasn't been updated
% to account for changes. See the new draft subsection on prevalence of recombination,
% though]

@@

% We present sc2ts, a real-time ARG inference method,
% and applied it to build a $\sim$2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
% By identifying specific samples as parents,
% sc2ts enables more precise interpretation of recombinants
% than state-of-the-art recombination detection methods,
% which find recombinants by matching a given sample
% to pre-defined population-averaged mutational profiles of Pango lineages.
% Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution,
% in contrast to RIPPLES \cite{Turakhia2022}
% which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
% inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution.
% Combined with new methods and tools to evaluate evidence for recombination,
% sc2ts is a significant technological advance over current methods
% for the discovery and analysis of recombinants and
% efficient inference of SARS-CoV-2 genealogy.

% The ARG itself is a useful resource to support future studies of SARS-CoV-2.
% Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
% which are shown in different studies
% using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
% The ARG displays remarkable phylogenetic congruence
% with a state-of-the-art phylogeny built using UShER.
% Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
% allowing for convenient and efficient reuse of this valuable dataset
% for a variety of evolutionary and genetic analyses beyond those presented here.
% % We hope that the ARG will spur on new research areas,
% % in which recombination is naturally and fully accounted for,
% % and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% % has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
% So far, recent developments in ARG methods have focused almost 
% exclusively on applications to human sequencing data.
% In that space, the availability of ARGs reconstructed from large datasets
% has already had tremendous impact on improving our understanding of
% human genetic ancestry and past demography \cite{Wohns2022}.
% We expect similar leaps to be enabled
% through tailoring ARG reconstruction methods
% to data from other species and pathogens.
% Specifically, existing inference methods can be adapted
% to use the sc2ts ARG for the genealogy-based study of selection \cite{hejase2022deep}
% or the geographic spread of the virus through time \cite{Grundler2025geographic}.
% This is in addition to simpler analyses,
% such as calculating mutation spectra
% or querying the number or frequency of mutations at a given position,
% which the ARG makes significantly more efficient, and
% ultimately more easily reproducible.
% Since the ARG is a condensed representation of the genetic information
% contained in the sequencing data, it can also be used to understand,
% for instance, how the variability in sequencing efforts across time 
% and location affects our ability to reconstruct the origin of 
% important lineages and detect new recombinants.
% Since reconstructing the ARG is computationally expensive, 
% and checking and improving various aspects of its quality is laborious, 
% we hope that making this resource openly available will save significant time
% and effort for other researchers and enable immediate focus on analyzing
% the information contained within this data. In-depth analysis of the
% ARG is also incredibly useful for identifying important directions for
% further research, both regarding methods (where current models do not
% correctly capture features of the data) and data processing 
% (where gaps in sequencing or data artifacts result in clear
% reconstruction errors).

% We contribute several novel methods and concepts
% which may find applications beyond sc2ts.
% First, we devise a simplified LS model that extends the classic concept of parsimony.
% In sc2ts, this model allows for accelerated HMM likelihood calculations,
% enabling reconstruction of the pandemic-scale ARG presented here.
% Existing HMMs used in statistical genetics may be modified
% in a similar way for significant speedup.
% Second, we propose the concept of mutations averted by recombination,
% which leads to an interpretable quantity to weigh the support for 
% the hypothesis that a given sample is a recombinant against
% the hypothesis that it is a non-recombinant,
% considering the sampled genetic diversity in an ARG and a specific HMM.
% Third, we propose the concept of HMM cost
% to identify poorly explained samples
% which are either too divergent or
% have a high number of mutations inconsistent with their collection dates.
% Here we use HMM cost to reject time-traveling samples,
% which no existing phylogenetic method detects and handles during inference
% (e.g., in the Viridian UShER phylogeny, parent and child nodes
% do not necessarily follow time order \cite{Hunt2024}),
% resulting in a cleaner dataset without samples
% with wrong collection dates that may affect downstream analyses.
% Future work may be to estimate dates
% for the suspected time traveling samples using the ARG provided here and
% the sc2ts sample matching engine.

% Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
% but has been understudied in part
% due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
% While the methods and tools developed here are tailored for SARS-CoV-2,
% they can be extended and adapted for other viruses.
% A major advantage of the sc2ts methods and tools is that
% they are built on top of the mature software libraries of tskit and VCF-Zarr,
% which enable pandemic-scale analyses that can be performed on typical laptops.
% Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
% making them highly adaptable for a broad range of applications.
% Together with the pandemic-scale ARG,
% we hope that sc2ts will be highly accessible to many researchers,
% enabling and accelerating studies of SARS-CoV-2 for years to come.

% \subsection*{Prevalance of recombination in SARS-CoV-2}
% [Rough first draft taken from earlier location. Sticking in a heading to
% make it clear it's distinct from other paragraphs.]

% The number of recombination events that we detect is largely 
% driven by the choice of recombination penalty, $k$ (or $p$
% for RIPPLES). 
% At $k=4$ we detect fewer recombinants overall than the combination of 
% UShER and RIPPLES at the same threshold. 
% However, most of the extra recombinants detected by UShER+RIPPLES
% involve violations of time ordering; whether this is due to incorrect
% metadata or these are false positives would require further study.
% However, at $p=4$, UShER+RIPPLES detects very few of the well-characterised
% Pango X-lineages which are detected by sc2ts. At the more lenient
% threshold of $p=3$, more of these events are detected (but still fewer
% than sc2ts) but at the cost of nearly 4 times more recombination
% events overall. Our detailed analysis shows that RIPPLES has a well
% defined task, which it performs well, finding the potential events
% in an exhaustive search. However, it can only operate within the
% input that it is provided, and it is the UShER tree provided with
% the Viridian data that does not have clear signals for recombination
% of X lineages such as XBB at the $p=4$ threshold. 
% The inference of recombination is therefore intrinsically linked
% with the underlying genealogy inference method.

% [Same point for us with XB. We don't pick up XB because of a long 
% branch and we don't have the right parents present.]

% [Whether k=3 or k=4 is ``correct'' is a matter for debate, and feeds
% into the next point.]

% The estimate for effective recombination in SARS-CoV-2 that we obtain
% at k=4 is much lower 
% than previous studies have suggbalanced
% % Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% % recombinant origins out of the first 87,695 genomes to be released
% In an early analysis, 225 potentially recombinant sequences out 
% of 87,695 samples collected before October 2020\cite{Varabyou2021rapid}.
% Over the same time period, XX events are present in the sc2ts ARG,
% representing a total of XX samples.
% In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
% 589 unique recombination events in a tree of 1.6 million samples
% up to May 2021 were identified after extensive quality control.
% These nodes had a combined total of 43,104 descendants, suggesting
% that approximately 2.7\% of genomes were of recombinant origin.
% Over the same time period, the sc2ts ARG contains 14 recombinants
% passing QC with a total of 172 descendants. The ARG contains a total
% 501,235 samples over this time period, giving an overall total fraction of
% 0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
% events and 79-fold descendant samples over the same time period.
% The numbers are not directly comparable as the published RIPPLES 
% estimate is based on GISAID data which is enriched for
% artefacts~\cite{Hunt2024}, but given the observations in 
% the previous section on running RIPPLES on the Viridian data
% it seems likely that RIPPLES is overestimating the prevalence
% of recombination in SARS-CoV-2.
